RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-10-23 doc/paper.tex
Local search finished. Control parameters described as well.

    \begin{algorithm}
    \small
    \hspace*{\algorithmicindent} \textbf{Input}: input training dataset $D_{tr}$ \\
    \hspace*{\algorithmicindent} \textbf{Output}: best symbolic formula solution $bs$
   	\begin{algorithmic}[1] 
   		\Procedure{RILS-ROLS}{$D_{tr}$}
   		\State $n_{tr} \gets |D_{tr}|$
   		\State $sample_{size} \gets $ \texttt{InitialSampleSize}($n_{tr}$)
   		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)
   		\State $s \gets NodeConstant(0)$ 
   		\State $s_{fit} \gets$ \texttt{Fitness}($s, D_{tr}'$)
   		\State $bs, bs_{fit} \gets s, s_{fit}$ \label{line:solSet}
   		\State $start_{tried}, perturbations_{tried} \gets \emptyset, \emptyset$
   		\While{\emph{stopping criteria is not met}}
   			\State $start_{tried} \gets start_{tried} \cup \{s\}$
   			\State $s_{perturbations} \gets $ \texttt{All1Perturbations}($s$) \label{line:sPert}
   			\State $s_{perturbations} \gets $ \texttt{FitOLS}($s_{perturbations}$, $D_{tr}'$)
   			\State $s_{perturbations} \gets $ \texttt{OrderByR2}($s_{perturbations}$) \label{line:orderR2}
   			\State $improved \gets false $
   			\For{$p \in s_{perturbations}$}
   				\If{$p \in perturbations_{tried}$}
   					\State \textbf{continue}
   				\EndIf
  				\State $p \gets $ \texttt{Simplify}($p$) \label{line:simp}
  				\State $p \gets $ \texttt{LocalSearch}($p$, $D_{tr}'$) \label{line:ls}
  				\State $p_{fit} \gets$ \texttt{Fitness}($p$, $D_{tr}'$)
  				\If{$p_{fit} < bs_{fit}$} \label{line:avoid1}
  					\State $bs, bs_{fit}, improved \gets p, p_{fit}, true$ // new best solution
  					\State \textbf{break}
  				\EndIf \label{line:avoid2}
  				\State $perturbations_{tried} \gets perturbations_{tried} \cup \{p\}$
   			\EndFor
   			\If{improved}
   				\State $s \gets bs$ \label{line:improved}
   			\Else
   				\State $start_{candidates} \gets $ \texttt{All1Perturbations}($bs$)
   				\If{$start_{candidates} \setminus start_{tried} = \emptyset$} // all 1-perturbations around $bs$ tried
   					\State $s' \gets $ \texttt{RandomPick}($start_{candidates}$)  \label{line:pert21}
   					\State $start_{candidates} \gets $ \texttt{All1Perturbations}($s'$) \label{line:pert22}
   				\EndIf
   				\State $s \gets $ \texttt{RandomPick}($start_{candidates} \setminus start_{tried}$) \label{line:randPick}
   				\If{not improved for too many iterations}
   					\State $sample_{size} \gets$ \texttt{IncreaseSampleSize}($sample_{size}, n_{tr}$) \label{line:sampleAdj1}
   					\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)\label{line:sampleAdj2}
   				\EndIf
   			\EndIf
   			\If{$R^2$ almost 1 and RMSE almost 0}
   				\State \textbf{break} // early exit
   			\EndIf
   		\EndWhile
   		\State \Return $bs$
   		\EndProcedure
   	\end{algorithmic}
   	\caption{RILS-ROLS method.}
   	\label{alg:rilsrols}
   \end{algorithm}

RILS-ROLS algorithms receives training dataset $D_{tr}$ as input. $D_{tr}$ can have very large number of records, so the first step of RILS-ROLS is to choose a random sample $D_{tr}' \subseteq D_{tr}$ of size $sample_{size}$. Initial the sample size is calculated as $\max(0.01 \cdot n_{tr}, 100)$. The size of sample is later dynamically adjusted through the algorithm's iterations -- when there are no solution improvements for some number of iterations, the sample size is doubled (lines~\ref{line:sampleAdj1}-\ref{line:sampleAdj2} of Algorithm~\ref{alg:rilsrols}).

@@

Main loop iterates as long as none of termination criteria is met: ($i$) maximal running time has been reached; ($ii$) maximal number of fitness calculations has been made; ($iii$) best solution is sufficiently good, 
w.r.t. the $R^2$ and $RMSE$ of the best solution. More precisely, if $R^2$ is sufficiently close to 1 and, at the same time, $RMSE$ is sufficiently close to 0, the algorithm stops prematurely -- this significantly reduces the running times for the majority of tested instances.

@@

@@

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local perturbations (1-perturbations) around solution $s$ -- $s_{perturbations}$
	\begin{algorithmic}[1] 
		\Procedure{All1Perturbations}{$s$}
		\State $s_{perturbations} \gets \emptyset$ \label{line:pertInit}
		\State $s \gets$ \texttt{NormalizeConstants}($s$)
		\State $s \gets$ \texttt{Simplify}($s$)
		\State $s_{subtrees} \gets$ \texttt{SubTrees}($s$)
		\For{$ss \in s_{subtrees}$} \label{line:forSSStart}
			\If{$ss = s$}
				\State $s_{perturbations} \gets s_{perturbations} \cup$ \texttt{NodeChange}($ss$)
			\EndIf
			\If{$ss.arity \geq 1$}
				\State $node_{perturbations} \gets$ \texttt{NodeChange}($ss.left$)
				\For{$np \in node_{perturbations}$}
					\State $ss_{new} \gets$ \texttt{Replace}($ss$, $ss.left$, $np$)
					\State $s_{perturbations} \gets s_{perturbations} \cup \{ss_{new}\}$
				\EndFor
			\EndIf
			\If{$ss.arity = 2$}
				\State $node_{perturbations} \gets$ \texttt{NodeChange}($ss.right$)
				\For{$np \in node_{perturbations}$}
					\State $ss_{new} \gets$ \texttt{Replace}($ss$, $ss.right$, $np$)
					\State $s_{perturbations} \gets s_{perturbations} \cup \{ss_{new}\}$
				\EndFor
			\EndIf
		\EndFor \label{line:forSSEnd}
		\State \Return $s_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of 1-perturbations.}
	\label{alg:pert}
\end{algorithm}

Initially, the set of perturbations $s_{perturbations}$ is empty (line~\ref{line:pertInit}).

@@

Perturbations are generated by making simple changes on the per-node level of $s$ expression tree.
Depending on the structure of the expression tree (expression does not have to have unique tree representation), the set of subtrees of the previous expression $(x+y^3)\cdot x/\sqrt{y}$ might be $\{(x+y^3)\cdot x/\sqrt{y}, (x+y^3), x/\sqrt{y}, x, y^3, \sqrt{y}, y\}$. This set of subtrees is obtained if left subtree of the whole expression is $(x+y^3)$ while the right one is $x/\sqrt{y}$. 
Lines \ref{line:forSSStart}-\ref{line:forSSEnd} show that there are three possibly overlapping cases when performing perturbations on the per-subtree level.

\emph{Case 1}. 
Subtree $ss$ is the whole tree $s$. Following on the the previous exemplary expression tree, this means that multiplication node that connects $(x+y^3)$ and $x/\sqrt{y}$ is to be changed. For example, multiplication can be replaced by addition, which forms a perturbation $(x+y^3)+ x/\sqrt{y}$.

\emph{Case 2}. 
Subtree $ss$ has arity of at least 1. This means that the left subtree exists, so the left subtree node is to be changed. 
For example, if $ss=x+y^3$ the overall perturbation might be $(x/y^3)+x/\sqrt{y}$ (addition is replaced by division). 
Another example would be the case of unary operation, e.g. when $ss=\sqrt{y}$. In that case, some of the possible perturbations could be $(x+y^3)\cdot x/\sqrt{\ln{y}}$ (application of logarithm to left subtree $y$) or $(x+y^3)\cdot x/\sqrt{x}$ (changing variable $y$ to $x$), etc.

@@

@@

\subsection{Parameter tuning}
SAMO JEDAN PARAMETAR!!!

>>>>

    \begin{algorithm}
    \footnotesize
    \hspace*{\algorithmicindent} \textbf{Input}: input training dataset $D_{tr}$ \\
     \hspace*{\algorithmicindent} \textbf{Control parameters}: size penalty $penalty_{size}$, error tolerance $tolerance_{error}$ \\
    \hspace*{\algorithmicindent} \textbf{Output}: best symbolic formula solution $bs$
   	\begin{algorithmic}[1] 
   		\Procedure{RILS-ROLS}{$D_{tr}$}
   		\State $n_{tr} \gets |D_{tr}|$
   		\State $sample_{size} \gets $ \texttt{InitialSampleSize}($n_{tr}$)
   		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)
   		\State $s \gets NodeConstant(0)$ 
   		\State $s_{fit} \gets$ \texttt{Fitness}($s, D_{tr}'$)
   		\State $bs, bs_{fit} \gets s, s_{fit}$ \label{line:solSet}
   		\State $start_{tried}, perturbations_{tried} \gets \emptyset, \emptyset$
   		\While{\emph{stopping criteria is not met}}
   			\State $start_{tried} \gets start_{tried} \cup \{s\}$
   			\State $s_{perturbations} \gets $ \texttt{All1Perturbations}($s$) \label{line:sPert}
   			\State $s_{perturbations} \gets $ \texttt{FitOLS}($s_{perturbations}$, $D_{tr}'$)
   			\State $s_{perturbations} \gets $ \texttt{OrderByR2}($s_{perturbations}$) \label{line:orderR2}
   			\State $improved \gets false $
   			\For{$p \in s_{perturbations}$}
   				\If{$p \in perturbations_{tried}$}
   					\State \textbf{continue}
   				\EndIf
  				\State $p \gets $ \texttt{Simplify}($p$) \label{line:simp}
  				\State $p \gets $ \texttt{LocalSearch}($p$, $D_{tr}'$) \label{line:ls}
  				\State $p_{fit} \gets$ \texttt{Fitness}($p$, $D_{tr}'$)
  				\If{$p_{fit} < bs_{fit}$} \label{line:avoid1}
  					\State $bs, bs_{fit}, improved \gets p, p_{fit}, true$ // new best solution
  					\State \textbf{break}
  				\EndIf \label{line:avoid2}
  				\State $perturbations_{tried} \gets perturbations_{tried} \cup \{p\}$
   			\EndFor
   			\If{improved}
   				\State $s \gets bs$ \label{line:improved}
   			\Else
   				\State $start_{candidates} \gets $ \texttt{All1Perturbations}($bs$)
   				\If{$start_{candidates} \setminus start_{tried} = \emptyset$} // all 1-perturbations around $bs$ tried
   					\State $s' \gets $ \texttt{RandomPick}($start_{candidates}$)  \label{line:pert21}
   					\State $start_{candidates} \gets $ \texttt{All1Perturbations}($s'$) \label{line:pert22}
   				\EndIf
   				\State $s \gets $ \texttt{RandomPick}($start_{candidates} \setminus start_{tried}$) \label{line:randPick}
   				\If{not improved for too many iterations}
   					\State $sample_{size} \gets$ \texttt{IncreaseSampleSize}($sample_{size}, n_{tr}$) \label{line:sampleAdj1}
   					\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)\label{line:sampleAdj2}
   				\EndIf
   			\EndIf
   			\If{$R^2$ almost 1 and RMSE almost 0 w.r.t. $tolerance_{error}$}
   				\State \textbf{break} // early exit
   			\EndIf
   		\EndWhile
   		\State $bs \gets $ \texttt{Simplify}($bs$)
   		\State $bs \gets $ \texttt{RoundModelCoefficients}($bs$)
   		\State \Return $bs$
   		\EndProcedure
   	\end{algorithmic}
   	\caption{RILS-ROLS method.}
   	\label{alg:rilsrols}
   \end{algorithm}

RILS-ROLS algorithms receives training dataset $D_{tr}$ as input. In addition, it has two control parameters: $penalty_{size}$ and $tolerance_{error}$. The first one quantifies the importance of solution expression complexity in the overall solution quality measure (more on this in Section~\ref{sec:fitness}). The second parameter is related to the expected noise level in data -- higher noise means that tolerance to errors should be higher.
 $D_{tr}$ can have very large number of records, so the first step of RILS-ROLS is to choose a random sample $D_{tr}' \subseteq D_{tr}$ of size $sample_{size}$. Initial the sample size is calculated as $\max(0.01 \cdot n_{tr}, 100)$. The size of sample is later dynamically adjusted through the algorithm's iterations -- when there are no solution improvements for some number of iterations, the sample size is doubled (lines~\ref{line:sampleAdj1}-\ref{line:sampleAdj2} of Algorithm~\ref{alg:rilsrols}).

@@

Main loop iterates as long as none of termination criteria is met: ($i$) maximal running time has been reached; ($ii$) maximal number of fitness calculations has been made; ($iii$) best solution is sufficiently good, 
w.r.t. the $R^2$ and $RMSE$ of the best solution. More precisely, if $R^2$ is sufficiently close to 1 and, at the same time, $RMSE$ is sufficiently close to 0, the algorithm stops prematurely -- this significantly reduces the running times for the majority of tested instances.
The sufficiency is controlled with parameter $tolerance_{error}$ -- smaller value means that $R^2$ and $RMSE$ should be closer to 1 and 0, respectively.

@@

Before returning the final symbolical model, RILS-ROLS peforms the final symbolical simplification and rounding of model coefficients. The later is sensitive to control parameter $tolerance_{error}$ -- smaller value means information loss during rounding is smaller, i.e. higher number of significant digits is preserved. Note that confidence in the symbolical model is reduced when expected noise in data is higher -- the rule of thumb is: for higher expected noise $tolerance_{error}$ should be higher.

@@

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local perturbations (1-perturbations) of solution $s$ -- $s_{perturbations}$
	\begin{algorithmic}[1] 
		\Procedure{All1Perturbations}{$s$}
		\State $s_{perturbations} \gets \emptyset$ \label{line:pertInit}
		\State $s \gets$ \texttt{NormalizeConstants}($s$)
		\State $s \gets$ \texttt{Simplify}($s$)
		\State $s_{subtrees} \gets$ \texttt{SubTrees}($s$)
		\For{$n \in s_{subtrees}$} \label{line:forSSStart}
			\State $n_{perturbations} \gets$ \texttt{All1PerturbationsAroundNode}($s$, $n$)
			\State $s_{perturbations} \gets s_{perturbations} \cup n_{perturbations}$	
		\EndFor \label{line:forSSEnd}
		\State \Return $s_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of all 1-perturbations for a given solution.}
	\label{alg:pert}
\end{algorithm}

Initially, the set of perturbations $s_{perturbations}$ is empty (line~\ref{line:pertInit} of Algorithm~\ref{alg:pert}).

@@

Perturbations are generated by making simple changes on the per-node level of $s$ expression tree.
Depending on the structure of the expression tree (expression does not have to have unique tree representation), the set of subtrees of the previous expression $(x+y^3)\cdot x/\sqrt{y}$ might be $\{(x+y^3)\cdot x/\sqrt{y}, (x+y^3), x/\sqrt{y}, x, y^3, \sqrt{y}, y\}$. This set of subtrees is obtained if left subtree of the whole expression is $(x+y^3)$ while the right one is $x/\sqrt{y}$. 
For each subtree $n$, the set of perturbations around $n$ is generated (lines \ref{line:forSSStart}-\ref{line:forSSEnd} of Algorithm~\ref{alg:pert}).

Algorithm~\ref{alg:pertNode} shows how perturbations around given subtree (node) are generated.

\begin{algorithm}	
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$, node $n$\\
	\hspace*{\algorithmicindent} \textbf{Output}: 1-perturbations of solution $s$ around node $n$
	\begin{algorithmic}[1]
		\Procedure{All1PerturbationsAroundNode}{$s$, $n$}
		\State $n_{perturbations} \gets \emptyset$
		\If{$n = s$}
		\State $n_{perturbations} \gets n_{perturbations} \cup$ \texttt{NodeChanges}($n$)
		\EndIf
		\If{$n.arity \geq 1$}
		\State $n_{changes} \gets$ \texttt{NodeChanges}($n.left$)
		\For{$nc \in n_{changes}$}
		\State $new \gets$ \texttt{Replace}($s$, $n.left$, $nc$)
		\State $n_{perturbations} \gets n_{perturbations} \cup \{new\}$
		\EndFor
		\EndIf
		\If{$n.arity = 2$}
		\State $n_{changes} \gets$ \texttt{NodeChanges}($n.right$)
		\For{$nc \in n_{changes}$}
		\State $new \gets$ \texttt{Replace}($s$, $n.right$, $nc$)
		\State $n_{perturbations} \gets n_{perturbations} \cup \{new\}$
		\EndFor
		\EndIf
		\State \Return $n_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of 1-perturbations for a given solution around given node.}
	\label{alg:pertNode}
\end{algorithm}

It can be seen that there are three possibly overlapping cases when performing perturbations on the per-node level.

\emph{Case 1}. 
Observed node $n$ is the whole tree $s$. Following on the the previous exemplary expression tree, this means that multiplication node that connects $(x+y^3)$ and $x/\sqrt{y}$ is to be changed. For example, multiplication can be replaced by addition, which forms a perturbation $(x+y^3)+ x/\sqrt{y}$.

\emph{Case 2}. 
Node $n$ has arity of at least 1. This means that the left subtree exists, so the left subtree node is to be changed. 
For example, if $n=x+y^3$ the overall perturbation might be $(x/y^3)+x/\sqrt{y}$ (addition is replaced by division). 
Another example would be the case of unary operation, e.g. when $n=\sqrt{y}$. In that case, some of the possible perturbations could be $(x+y^3)\cdot x/\sqrt{\ln{y}}$ (application of logarithm to left subtree $y$) or $(x+y^3)\cdot x/\sqrt{x}$ (changing variable $y$ to $x$), etc.

@@

Perturbations are further improved by means of local search procedure (Algorithm~\ref{alg:ls}).

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: perturbation $p$, sample training dataset $D_{tr}'$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local optimum $bp$ in the vicinity of perturbation $p$
	\begin{algorithmic}[1] 
		\Procedure{LocalSearch}{$p$, $D_{tr}'$}
		\State $bp, bp_{fit} \gets p, $\texttt{Fitness}$(p, D_{tr}')$ 
		\State $improved \gets true$
		\While{$improved$}
			\State $improved \gets false$
			\State $bp_{candidates} \gets \emptyset$
			\State $bp_{subtrees} \gets $\texttt{SubTrees}($bp$)
			\For{$n \in bp_{subtrees}$}
				\State $n_{candidates} \gets $ \texttt{All1PerturbationsAroundNode}($bp$, $n$)
				\For{$new \in n_{candidates}$}
					\State $new \gets$ \texttt{FitOLS}($new$, $D_{tr}'$)
					\State $new_{fit} \gets$ \texttt{Fitness}($new$, $D_{tr}'$)
					\If{$new_{fit} < bp_{fit}$}
						\State $bp, bp_{fit} \gets new, new_{fit}$
						\State $improved \gets true$
					\EndIf
				\EndFor
			\EndFor
		\EndWhile
		\State \Return $bp$
		\EndProcedure
	\end{algorithmic}
	\caption{Local search procedure.}
	\label{alg:ls}
\end{algorithm}

For a given perturbation $p$, local search systematically explores all 1-perturbations around $p$. It relies on the \emph{best-improvement} strategy, meaning that all 1-perturbations (for all subtrees) are considered. Before checking if the candidate solution ($new$) is better that the actual best $bp$, OLS coefficient fitting is performed.

@@

\subsection{Parameter tuning}
SAMO JEDAN PARAMETAR!!!
Zapravo dva parametra $penalty_{size}$ i $tolerance_{error}$ pri cemu je drugi manje vise zavistan od ocekivanog 
nivoa suma -- ne znam kako obrazloziti njegov tuning (mozda prema greski na trening skupu pa izabrati onaj za koji su R2 i RMSE bolji...).
